% =============================================================================
% Chapter 7: Non-Functional Requirements
% =============================================================================

\chapter{Non-Functional Requirements}
\label{ch:nonfunctional}

\section{Reliability and Testing}

\begin{reqbox}{REQ-74}{\stable}
\textbf{Test-driven development.}\\
Every public method has its test written before its implementation. Pull
requests adding or modifying public API without a corresponding test are
rejected.
\end{reqbox}

\begin{reqbox}{REQ-75}{\stable}
\textbf{Coverage threshold.}\\
Test coverage measured with \code{pytest-cov} does not fall below 90\,\%
on the main branch. This is a hard CI gate, not an aspirational target.
\end{reqbox}

\begin{reqbox}{REQ-76}{\stable}
\textbf{Required edge-case tests.}\\
The test suite includes the cases enumerated in
Table~\ref{tab:required-edge-cases}, organized by family for audit ease.
\end{reqbox}

\begin{table}[H]
\centering
\caption{Required edge-case tests, grouped by area.}
\label{tab:required-edge-cases}
\small
\begin{tabularx}{\textwidth}{@{}lX@{}}
\toprule
\textbf{Area} & \textbf{Edge case}\\
\midrule
Load        & Empty VarFrame; single source file; dict-mode coordinate-tuple length mismatch; pandas DataFrame with non-string column names; NumPy array with mismatched \code{names}.\\
Pivot       & Datapoint mode with no subsequent pivot; \code{pivot()} on already-structured VarFrame; \code{auto\_detect=True} on ambiguous candidates; duplicate coordinates on every requested dimension raise \code{PivotDuplicateError}.\\
Diagnostics & 100\,\%-coverage VarFrame; all-NaN slice; single-point dimension.\\
Compute     & Invalid RPN expression (syntax error, undefined variable); cyclic \code{.itceq} equations caught at parse time; non-differentiable NumPy with uncertainty guard (per-variable); \code{where} with \code{fill=None} on non-existent variable; \code{compute} with both symbolic and Monte Carlo methods on the same expression.\\
Ops         & \code{diff} with \code{window <= deg}; \code{interpolate} with non-monotonic axis-translation target; \code{interpolate} with \code{override=False} on existing coordinate; \code{expand} with duplicate existing dimension name; \code{itc.concat} with overlapping \code{along} values; \code{fitvalue} evaluation outside the original fit range producing \code{-1} tags; \code{combine} with mismatched grids; numerical operation on non-numeric dimension.\\
Compare     & Grid mismatch in \code{compare} without reference; multiple named inputs with overlapping deltas.\\
Integrate   & \code{np.nan} inside the integration domain with \code{skipna=False} (result NaN) and \code{skipna=True} (populated cells only, History records the fraction).\\
Immutability & In-place mutation of any stored array raises; \code{to\_numpy} returns a read-only view by default.\\
Reproducibility & Two identical load-and-operate runs yield identical state hashes; hashes are invariant to timestamps and user identity.\\
History     & Hash mismatch on reload; pipeline applied to incompatible VarFrame; comment preserved through \code{.itc} round-trip.\\
Modes       & Draft-mode export blocked without flag; mixed-mode binary operation rejected; \code{db.promote} / \code{db.demote} round trip preserves data.\\
Uncertainty & Mismatched uncertainty keys; correlation matrix non-symmetric; correlation matrix with $|r| > 1$; declared correlation between non-existent variables; repeat averaging reduces the random component by $1/\sqrt{N}$ and leaves the systematic component unchanged.\\
Axes        & \code{rotate} with unregistered axis; vector group cannot be resolved; non-orthogonal rotation matrix rejected; uncertainty propagated through rotation matches direct LPU.\\
\bottomrule
\end{tabularx}
\end{table}

\begin{reqbox}{REQ-77}{\stable}
\textbf{Property-based tests for math kernels.}\\
Property-based tests using Hypothesis~\citep{hypothesis} are required for:
\begin{itemize}[noitemsep,topsep=2pt]
  \item all RPN operators (verifying analytical partial-derivative
    identities on randomized inputs);
  \item the propagation engine (verifying additivity of variances under
    independent inputs, correctness under known correlated inputs, and
    dimensional consistency);
  \item the \code{diff} kernel (verifying exact recovery of polynomials
    of degree $\le \code{deg}$);
  \item the rotation engine (verifying that
    $R^\top R = I$ is preserved, and that uncertainty propagated through
    rotation matches direct LPU computation).
\end{itemize}
\end{reqbox}

\section{Code Quality}

\begin{reqbox}{REQ-78}{\stable}
\textbf{Type hints.}\\
All public functions and methods carry full Python type hints.
\code{mypy --strict} is run in CI on the public API.
\end{reqbox}

\begin{reqbox}{REQ-79}{\stable}
\textbf{Docstrings.}\\
All public functions and classes carry NumPy-format docstrings with
Parameters, Returns, Raises, and at least one Examples section.
\end{reqbox}

\begin{reqbox}{REQ-80}{\stable}
\textbf{Linting and formatting.}\\
All code passes \code{ruff check} linting and \code{ruff format}
formatting. One tool covers both concerns. CI runs both checks on every
push and rejects non-conforming code.
\end{reqbox}

\begin{reqbox}{REQ-81}{\stable}
\textbf{Error messages.}\\
All exception messages include three pieces of information: the variable
or object involved; the operation attempted; and a suggested fix. Errors
derive from the \code{ITACAError} hierarchy
(Table~\ref{tab:exception-hierarchy}).
\end{reqbox}

\section{Architecture and Dependencies}

\begin{reqbox}{REQ-82}{\stable}
\textbf{NumPy-only core.}\\
The \code{core/}, \code{ops/}, and \code{uncertainty/} packages depend
only on NumPy. No xarray, dask, or pandas imports appear in these
modules. CI enforces this with a \code{ruff} import-policy rule.
\end{reqbox}

\begin{reqbox}{REQ-83}{\stable}
\textbf{Pinned dependency version ranges.}\\
\itc{} declares explicit version ranges for every dependency in
\code{pyproject.toml}. Table~\ref{tab:dependency-versions} lists the
minimum tested version, the upper bound, and whether the dependency is
required, optional, or development-only. CI runs the full test suite
against the minimum and the latest compatible versions.
\end{reqbox}

\begin{table}[H]
\centering
\caption{Dependency version policy for \itc{} v0.1.0. Bounds are
expressed as PEP 440 specifiers.}
\label{tab:dependency-versions}
\small
\begin{tabularx}{\textwidth}{@{}llXl@{}}
\toprule
\textbf{Dependency} & \textbf{Role} & \textbf{Used for} & \textbf{Range}\\
\midrule
\code{numpy}        & required  & Core arrays, RPN, propagation, ops          & \code{>=1.26,<3.0}\\
\code{python}       & required  & Language baseline                            & \code{>=3.10,<3.14}\\
\code{matplotlib}   & optional  & Default plotting backend                     & \code{>=3.8,<4.0}\\
\code{plotly}       & optional  & Interactive plotting backend                 & \code{>=5.20,<6.0}\\
\code{pandas}       & optional  & DataFrame bridge                             & \code{>=2.1,<3.0}\\
\code{smt}          & optional  & Surrogate modeling                           & \code{>=2.5,<3.0}\\
\code{tectonic}     & optional  & LaTeX-backed PDF reports (preferred)         & system\\
\code{pdflatex}     & optional  & LaTeX-backed PDF reports (fallback)          & system\\
\code{pytest}       & dev       & Test runner                                  & \code{>=8.0,<9.0}\\
\code{pytest-cov}   & dev       & Coverage gate                                & \code{>=5.0,<6.0}\\
\code{hypothesis}   & dev       & Property-based testing                       & \code{>=6.100,<7.0}\\
\code{ruff}         & dev       & Linting and formatting                       & \code{>=0.5,<1.0}\\
\code{mypy}         & dev       & Static type checking                         & \code{>=1.10,<2.0}\\
\code{pre-commit}   & dev       & Local CI mirror                              & \code{>=3.7,<5.0}\\
\bottomrule
\end{tabularx}
\end{table}

\begin{reqbox}{REQ-84}{\stable}
\textbf{Optional dependency management.}\\
Optional dependencies (matplotlib, plotly, SMT, pandas, the LaTeX
toolchain) are imported lazily and checked at the point of use. A missing
optional dependency raises \code{MissingDependencyError} with an
actionable installation instruction (e.g.\
\code{pip install itaca[plot]}).
\end{reqbox}

\begin{reqbox}{REQ-85}{\stable}
\textbf{API parameter order.}\\
Canonical parameter order: \emph{(what, where, how)}: the variable or
object first, then location or dimension, then method or options.
Optional parameters beyond the first positional argument are
keyword-only.
\end{reqbox}

\begin{reqbox}{REQ-86}{\stable}
\textbf{English everywhere.}\\
All code, comments, docstrings, variable names, error messages,
\code{.itceq} files, and documentation are in English without exception.
\end{reqbox}

\begin{reqbox}{REQ-87}{\stable}
\textbf{American English with Z.}\\
\itc{} uses American English orthography consistently: \emph{organize}
(not organise), \emph{visualization} (not visualisation),
\emph{normalize} (not normalise). This reduces inconsistency and matches
the dominant convention in scientific Python documentation.
\end{reqbox}

\section{Performance and Memory}

\begin{reqbox}{REQ-88}{\stable}
\textbf{Vectorized operations.}\\
All operations on VarFrame arrays are vectorized via NumPy. Python loops
over array elements are not acceptable in performance-critical paths
(\code{compute}, propagation, \code{diff}, \code{smooth},
\code{interpolate}, \code{rotate}).
\end{reqbox}

\begin{reqbox}{REQ-89}{\stable}
\textbf{Memory transparency.}\\
\code{db.summary()} reports the in-memory footprint of a VarFrame.
Operations that potentially trigger large allocations (e.g.\ \code{expand}
broadcasting to a high-cardinality new dimension, dense pivot of a sparse
acquisition matrix) document the resulting size in their docstring and
print a warning when the projected allocation exceeds 1\,GB.
\end{reqbox}

\begin{reqbox}{REQ-90}{\stable}
\textbf{Sparse-aware handling of large datasets.}\\
For VarFrame instances whose coverage drops below a configurable
threshold (default 25\,\%), \itc{} provides sparse-aware variants of the
heavy operations (\code{compute}, \code{interpolate}, \code{integrate})
that operate only on populated cells. The sparse mode is opt-in via
\code{db.use\_sparse=True}; correctness must match the dense path
exactly, validated by paired tests. A common backing store (the underlying NumPy arrays plus a finite-mask) is used; \itc{} does not
introduce a separate sparse VarFrame type.
\end{reqbox}

\begin{reqbox}{REQ-91}{\stable}
\textbf{Lazy materialization of derived structures.}\\
The HistoryFrame, UncFrame, and CorrelationMatrix are materialized only
when first needed. A VarFrame loaded from raw data carries no UncFrame
until \code{set\_uncertainty} is called; no HistoryFrame until an
operation produces non-original values; no correlation matrix until
\code{set\_correlation} is called. This keeps the memory footprint of
exploratory workflows minimal.
\end{reqbox}

\section{Versioning and Release}

\begin{reqbox}{REQ-92}{\stable}
\textbf{Semantic versioning.}\\
\itc{} follows semantic versioning (\code{MAJOR.MINOR.PATCH}). Initial
public release is \code{0.1.0}. Breaking API changes increment
\code{MAJOR}; new backward-compatible features increment \code{MINOR};
bug fixes increment \code{PATCH}.
\end{reqbox}

\begin{reqbox}{REQ-93}{\stable}
\textbf{Conventional commits.}\\
All commit messages follow Conventional Commits. Allowed prefixes:
\code{feat:}, \code{fix:}, \code{docs:}, \code{test:}, \code{refactor:},
\code{chore:}, \code{perf:}, \code{build:}, \code{ci:}. CI rejects
non-conforming commits.
\end{reqbox}

\begin{reqbox}{REQ-94}{\stable}
\textbf{Changelog.}\\
\code{CHANGELOG.md} is updated with every version bump, following the
Keep a Changelog format~\citep{keepachangelog} (Added, Changed,
Deprecated, Removed, Fixed, Security).
\end{reqbox}

\begin{reqbox}{REQ-95}{\stable}
\textbf{Continuous integration.}\\
GitHub Actions runs on every push: \code{ruff check},
\code{ruff format --check}, \code{mypy --strict} (public API),
\code{pytest}, and the coverage gate (threshold 90\,\%). All checks must
pass before a pull request may be merged.
\end{reqbox}

\begin{reqbox}{REQ-96}{\stable}
\textbf{Pre-commit hooks aligned with CI.}\\
A \code{.pre-commit-config.yaml} mirrors the CI checks (\code{ruff check},
\code{ruff format --check}, \code{mypy}, fast \code{pytest -x}), running
locally on every commit. The local pre-commit configuration is a hard
requirement of the contribution workflow.
\end{reqbox}

\begin{reqbox}{REQ-97}{\draft}
\textbf{Citable releases.}\\
Tagged releases on GitHub are mirrored on Zenodo, generating a DOI for
each release. \itc{} is academically citable from \code{0.1.0}.
\end{reqbox}

\begin{reqbox}{REQ-108}{\draft}
\textbf{Published documentation site.}\\
\itc{} publishes a documentation site on GitHub Pages, built with
ProperDocs (the maintained MkDocs fork) and the material theme,
mirroring the pyflightstream site (DD-23). The navigation covers a
getting-started home page, the SRS chapters authored in markdown for
the site, \code{DECISIONS.md}, \code{OPEN\_QUESTIONS.md}, the
sister-library page, and an API reference. Following the pyflightstream
scheme, the reference and architecture pages are generated at build
time by \code{mkdocs-gen-files} and \code{mkdocs-literate-nav} from a
docs-build script that reads the live NumPy-format docstrings through
standard-library introspection, one page per public module and kept in
step with the milestone. The generator script and the mkdocs tooling
live in the docs build tree, are never imported by library code, and
nothing generated is committed, so the site cannot drift from the code
and the NumPy-only rule (REQ-82) is untouched. The renderer is a build
script rather than \code{mkdocstrings}, which pyflightstream dropped. A
\code{docs.yml} workflow builds the site with \code{--strict} on push
to \code{main} and deploys it to GitHub Pages. \code{properdocs},
\code{mkdocs-material}, \code{mkdocs-gen-files}, and
\code{mkdocs-literate-nav} are development-only dependencies, added to
the REQ-83 dependency table with version ranges and never runtime. The
published site does not replace the SRS as the authoritative
specification.
\end{reqbox}
