% =============================================================================
% Chapter 11: Changelog
% =============================================================================

\chapter{Changelog}
\label{ch:changelog}

This changelog follows the Keep a Changelog format~\citep{keepachangelog}.
Versions are listed in reverse chronological order. This chapter tracks the
\emph{document} baseline; library releases carry their own
\code{CHANGELOG.md} in the repository (REQ-94). From the next document
version onward, every entry lists the requirement(s) and design decision(s)
it adds or modifies.

\section*{Document 0.2.0, 2026-07-23, M1 Re-Baseline and Library Infrastructure}

Issued at the M1 kickoff (ultraplan Batch A approval), together with
the approved execution plan committed as
\code{docs/M1\_EXECUTION\_PLAN.md}.

\paragraph{Added.} Functional requirements REQ-104 (options registry
with exact keys), REQ-105 (typed no-default sentinel), and REQ-106
(accessor registration), all draft, adopted from the 2026-07-23
library-review triage (decisions D1, D2, D6) in the new
Section~\ref{sec:library-infrastructure}. Design-decision boxes DD-23
(co-development with pyflightstream, recorded in the decision log on
2026-07-23 and entered here at this baseline per the editing rule),
DD-24 (options-registry adoption), and DD-25 (\code{uncertainties} as
a dev-only test oracle, never a runtime dependency) in
Chapter~\ref{ch:architecture}, each with its long-form entry in the
repository decision log.

\paragraph{Changed.} Chapter~\ref{ch:roadmap}: the M1 deliverables
table splits into v0.2.0 (computation: operations, axes, pipeline,
processor infrastructure, REQ-105, REQ-106) and v0.2.1 (presentation
and builtins: REQ-104, plot core, WT builtins, statistics and
compare), recording the 2026-07-23 re-baseline decision; the stretch
scope may still ship inside v0.2.0 if green in the same window.
REQ-101 was validated as written at the Batch A checkpoint; the
reqbox stays draft until the implementation and its tests land, then
moves to stable through this changelog and the revision history.
At the M1 Phase B1 checkpoint, REQ-31 gained a normative note that
\code{fitmodel} raises when uncertainty is present (its
coefficient-space rule is not yet frozen, OQ-24), and the
Table~\ref{tab:unc-ops} normative UncFrame semantics gained a
\code{fitmodel}/\code{fitvalue} provisional row and a scoped
restatement of the linearity claim (weight-based rows exact, Jacobian
rows linearized, provisional rows raising). OQ-24 (coefficient-space
uncertainty) and OQ-25 (origin-tag reduction across collapsing and
expanding operations) were opened during the same phase. Three
author-seat calls were resolved at the B1 checkpoint: \code{fitvalue}
defers with \code{fitmodel} and raises on uncertainty (REQ-32,
OQ-24); the exception hierarchy replaces the \code{diff}-only
\code{DiffWindowError} with a shared \code{FitDegreeError} for the
too-few-points-for-degree invariant across \code{diff}, \code{smooth},
\code{interpolate}, and \code{fitmodel} (REQ-30,
Table~\ref{tab:exception-hierarchy}); and \code{fill}'s positional
\code{method} argument is deprecated toward keyword-only for
consistency with the M1 kernel operations (REQ-26).

\paragraph{Added (B2).} Functional requirement REQ-107 (per-group
source frames and multi-frame data), draft. Design decisions DD-26
(\code{scipy} as a dev-only direction-cosine test oracle) and DD-27
(the sanctioned accessor per-instance cache write) in
Chapter~\ref{ch:architecture}. The axis registry and vector-group
declarations join the state hash (REQ-103). The built-in wind and
stability frames follow the AIAA R-004A Etkin direction-cosine
convention, SME-accepted at the B2 checkpoint. The exception
hierarchy gains \code{AxisNotFoundError}, \code{VectorGroupError},
\code{RotationMatrixError}, and \code{AccessorRegistrationError}.

\paragraph{Changed (B2).} REQ-101 (condition-dependent axes) promoted
from draft to stable: condition-dependent frames are implemented and
tested (per-grid-point evaluation, chain-rule angle uncertainty,
proper-rotation validation, angle unit from the Dimension or Variable
metadata). The rotation propagation treats frame angles as mutually
independent, rejecting a declared correlation that involves an angle
rather than dropping it; the modeling question is recorded as OQ-26.
Three B2 author-seat calls were resolved (queue Q-013): \code{Axis}
is exported at the package top level; the axes surface standardizes on
``axis'' for a coordinate system (REQ-107 \code{declare\_vector(axis=)},
distinct from the \code{Frame=} selector of \code{select}); and
\code{translate\_moments} gains \code{force} and \code{moment} group
selectors (REQ-100). OQ-25 (origin-tag reduction across collapsing and
reshaping operations) was resolved and its four rules folded into the
HistoryFrame section (Chapter~\ref{ch:datamodel}, Section 4.3).

\paragraph{Changed (B3a).} The \code{.itc\_pipe} format
(Chapter~\ref{ch:datamodel}, Section~\ref{sec:itc-pipe}) is specified as JSON rather
than TOML, with design decision DD-28 recording why (no
standard-library TOML writer in any Python version, no TOML null type
against a meaningful \code{compute(fill=None)}, and nested replay
arguments) and the reasoning that every required content item is kept:
the creating version, the source index range, each call with its
keyword arguments and attached comment (REQ-19), and a content hash
that is reverified on load. REQ-53 gained the normative extraction
contract: frame construction (\code{load}, \code{pivot}) is input
preparation and is skipped when it leads the range, any other
operation that records no replayable step raises
\code{PipelineCompatibilityError}, and a range yielding no step raises
rather than returning a pipeline that would apply as a silent no-op.
REQ-54 gained the replay contract: the recorded call and comment
reproduce the state hash, and an incompatible target raises with the
underlying error's object and suggested fix preserved. The History-entry
table gained the \code{step} field (replay metadata, excluded from the
state hash, REQ-103 scope unchanged), and the \code{.itc} archive
persists it at schema \code{itaca-itc/2} so a reopened archive can still
lift its recipe. The archive section (Section~\ref{sec:itc-archive}) was
amended to match, naming both schema versions, the \code{step} member of
\code{history.json}, and the \code{steps\_hash} digest in
\code{metadata.json}. That digest covers the recipe while the REQ-103
state hash covers the recovered data: without it an edited step passes
the state check, because the stored data still matches itself, and the
edited recipe then steers the next replay. DD-28 also records the replay
representation itself:
structured step re-dispatch rather than re-parsing the History display
strings.

\paragraph{Added (docs adoption).} Non-functional requirement REQ-108
(published documentation site on GitHub Pages, ProperDocs with markdown
SRS chapters and a build-time-generated API reference, mirroring the
pyflightstream site), draft, in Chapter~\ref{ch:nonfunctional}. A
pyflightstream adoption (DD-23); recorded in
\code{PYFLIGHTSTREAM\_ADOPTIONS.md}.

\section*{Document 0.1.1, 2026-07-21, M0 Uncertainty-Semantics Validation}

Issued at the M0 Phase 4 checkpoint, with the implementation and its
test suite as evidence.

\paragraph{Changed.} REQ-98 and REQ-99 promoted from draft to stable;
the smooth and diff row of REQ-98 stays provisional until OQ-18 is
resolved during v0.2.0 implementation, and operations without a frozen
weight rule raise when uncertainty is present. REQ-01 and REQ-03 gain
the dict-mode \code{"*"} swept sentinel (OQ-19). REQ-01 restricts
string-valued columns to dimension coordinates (OQ-21). REQ-11 states
the draft-guard scope: result exports guarded, inspection artifacts
exempt (OQ-22). Section~4.2 states that declared correlation applies
to both uncertainty components (OQ-23).

\paragraph{Added.} The \code{source\_coords} field of the
\code{Provenance} object, backing \code{db.manifest} (OQ-20).

\section*{Document 0.1.0, 2026-07-21, First Workspace-Tracked Version}

This version establishes the full requirement and design baseline of
\itc{} and starts requirement-evolution monitoring in the research
workspace. It consolidates the founding v0.1.0b draft with the 2026-07-21
specification review.

\paragraph{Added.} Functional requirements REQ-01 through REQ-73 and
REQ-98 through REQ-103; non-functional requirements REQ-74 through REQ-97;
non-requirements NREQ-01 through NREQ-10; design decisions DD-01 through
DD-22. The \code{.itc}, \code{.itceq}, \code{.itc\_pipe}, and
\code{.itc\_surr} file formats. The positioning statement among existing
tools (Section~\ref{sec:positioning}). The MIT license and citation policy
(Chapter~\ref{ch:contributing}). The incremental release roadmap mapping
milestones M0 through M3 to public releases \code{v0.1.0} through
\code{v0.4.0} (Chapter~\ref{ch:roadmap}).

\paragraph{Changed relative to the founding draft.} Two-component
uncertainty replaces the single-component UncFrame (REQ-99, DD-19);
uncertainty semantics are now normative for every operation (REQ-98,
DD-18); the axis system gains condition-dependent frames and moment
reference-point transfer (REQ-100, REQ-101); array-level immutability and
the state-hash scope are now requirements (REQ-102, REQ-103); expression
parsing is specified on the standard-library \code{ast} module (DD-20);
\code{ruff format} replaces \code{black}; the Python upper bound moves to
\code{<3.14}; the single-release plan is superseded by incremental
releases (DD-21).
