% =============================================================================
% Revision History
% =============================================================================

\section*{Revision History}
\addcontentsline{toc}{section}{Revision History}

\begin{tabularx}{\textwidth}{@{}llXX@{}}
\toprule
\textbf{Version} & \textbf{Date} & \textbf{Author} & \textbf{Summary}\\
\midrule
0.2.0 & 2026-07-23 & Geovana Neves &
M1 re-baseline (ultraplan Batch A): the Chapter 10 M1 deliverables
split into a v0.2.0 computation table and a v0.2.1 presentation and
builtins table, with the stretch scope allowed to ship inside v0.2.0
if green in the same window. Library-infrastructure requirements
added as draft from the 2026-07-23 library-review triage: options
registry with exact keys (REQ-104, DD-24), typed no-default sentinel
(REQ-105), accessor registration (REQ-106). Design-decision boxes
DD-23 to DD-25 enter Chapter~5 with their long-form entries in the
decision log; DD-25 records the dev-only \code{uncertainties} test
oracle (no requirement). REQ-101 was validated as written at the same
checkpoint; its status moves to stable through the normal process
once implemented and tested during M1. At the M1 Phase B1
checkpoint, REQ-31 and the normative UncFrame-semantics table gained
the \code{fitmodel} raise-until-frozen note and row (OQ-24), and
OQ-24 and OQ-25 were opened. Three author-seat calls at the same
checkpoint: \code{fitvalue} defers with \code{fitmodel} and raises on
uncertainty (OQ-24); the too-few-points-for-degree invariant is
unified into a shared \code{FitDegreeError} leaf across \code{diff},
\code{smooth}, \code{interpolate}, and \code{fitmodel} (REQ-30, the
former \code{DiffWindowError}); and \code{fill}'s positional
\code{method} is deprecated toward keyword-only (REQ-26). These are
same-baseline refinements recorded here rather than a further version
bump. M1 Phase B2 opened the axes work: the built-in wind and
stability direction-cosine convention (AIAA R-004A, Etkin) was
SME-accepted; scipy joined the dev-only test oracles (DD-26);
condition-dependent angle inputs are read in the unit of their
Dimension or Variable; the axis registry and vector-group declarations
join the VarFrame state hash (REQ-103); and REQ-107 (per-group source
frames and multi-frame data) was added as a draft requirement. With the
axes machinery implemented and tested (rotation with per-grid-point
condition-dependent frames and chain-rule angle uncertainty, moment
transfer), REQ-101 was promoted from draft to stable. The angle
independence model of the rotation propagation was recorded as OQ-26,
and DD-27 records the sanctioned accessor per-instance cache write.
OQ-25 (origin-tag reduction across collapsing and reshaping
operations) was resolved and its four rules folded into Section 4.3.
REQ-108 (published documentation site on GitHub Pages, mirroring
pyflightstream, DD-23) was added as a draft requirement. M1 Phase B3a
implemented the pipeline requirements and corrected the specification
where it was wrong: Section~\ref{sec:itc-pipe} now specifies the
\code{.itc\_pipe} serialization as JSON rather than TOML, with the reasons recorded in
DD-28 (no standard-library TOML writer in any Python version, no TOML
null type against a meaningful \code{compute(fill=None)}, and nested
replay arguments); every content item the section required is kept,
including the attached comment (REQ-19) and the content hash, which is
now reverified on load. REQ-53 gained the normative extraction contract
(frame construction is skipped as input preparation, any other
step-less operation raises, and a range yielding no step raises rather
than producing a silent no-op), and REQ-54 gained the replay contract
(the recorded call and comment reproduce the state hash, and an
incompatible target raises with the underlying object and fix
preserved). The History-entry table gained the \code{step} field, replay
metadata excluded from the state hash and persisted in the \code{.itc}
archive at schema \code{itaca-itc/2}. Section~\ref{sec:itc-archive} was
amended to match: it now names both schema versions, the \code{step}
member of \code{history.json}, and the \code{steps\_hash} digest in
\code{metadata.json}, which covers the recipe while the REQ-103 state
hash covers the recovered data. Without the second digest an edited
step passes the state check and then steers the next replay. The
scope of REQ-103 is unchanged.\\
\addlinespace
0.1.1 & 2026-07-21 & Geovana Neves &
REQ-98 and REQ-99 promoted from draft to stable after the M0
implementation checkpoint: propagation semantics validated for every
M0 operation, with the smooth and diff row of REQ-98 remaining
provisional pending OQ-18. Declared correlation applies to both
uncertainty components (Section 4.2, OQ-23). Folded the OQ-19 to
OQ-22 resolutions into the text: dict-mode \code{"*"} swept sentinel
(REQ-01, REQ-03), Provenance \code{source\_coords} backing the
manifest (Section 4.4.1), string columns as dimension coordinates
only (REQ-01), and the draft-guard scope (REQ-11).\\
\addlinespace
0.1.0 & 2026-07-21 & Geovana Neves &
First workspace-tracked version. Consolidates the founding v0.1.0b draft and
incorporates the 2026-07-21 specification review: uncertainty semantics for
every operation (REQ-98), two-component uncertainty (REQ-99), moment
reference-point transfer (REQ-100), condition-dependent axes (REQ-101),
array immutability (REQ-102), state-hash reproducibility contract (REQ-103),
solver-automation boundary (NREQ-10), design decisions DD-14 to DD-22,
positioning among existing tools, license declaration, and an incremental
release roadmap. Requirement evolution is monitored from this baseline
onward: every change is recorded in this table and in
Chapter~\ref{ch:changelog}, and deprecated requirements are kept in place
and tagged \deprecated{}.\\
\bottomrule
\end{tabularx}

\vspace{1em}
\begin{noteinfo}[About this document]
This Software Requirements Specification is part of the \itc{} project's
public design record. It is being prepared in parallel with the implementation,
which is itself developed with assistance from large language model tooling
(notably Claude Code). Every implementation derived from these requirements
goes through human double-check by the author before being merged. The SRS is
the authoritative reference; if a discrepancy with the code is detected, the
code is corrected, not the SRS.
\end{noteinfo}

\clearpage
